The \proglang{R} Package \pkg{forestinventory}: Design-Based Global and Small Area Estimations for Multi-Phase Forest Inventories

### 2017-10-04 JStatSoft_forestinventory_master-concordance.tex
update Andreas 04.10.2017(1)

\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:%
1 107 1 1 4 1 1 1 4 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Introduction.Rnw:ofs 115:%
1 97 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 213:%
123 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Main structure and modules of the Package.Rnw:ofs 223:%
1 134 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 358:%
134 11 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Twophase_Estimators and their Application.Rnw:ofs 370:%
1 172 1 1 4 1 1 1 2 1 0 2 1 3 0 1 2 2 1 1 3 14 0 1 2 16 1 1 5 7 0 1 2 4 1 1 2 24 0 1 2 6 1 1 2 4 0 1 2 4 1 1 2 1 0 1 5 6 0 1 2 3 1 1 2 8 0 1 2 133 1 1 7 6 0 1 1 26 0 1 2 3 1 1 2 13 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 7 1 1 6 5 0 1 1 8 0 1 2 7 1 1 2 4 0 1 2 1 1 1 6 5 0 1 1 3 0 1 2 1 1 1 2 11 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 955:%
147}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Threephase_Estimators and their Application.Rnw:ofs 956:%
1 105 1 1 2 1 0 1 1 3 0 1 2 4 1 1 4 1 1 1 6 8 0 1 2 2 1 1 2 27 0 1 2 175 1 1 3 2 0 1 1 3 0 1 2 4 1 1 8 10 0 1 2 1 1 1 4 1 1 1 2 12 0 1 2 4 1 1 7 9 0 1 2 2 1 1 2 12 0 1 2 7 1 1 7 9 0 1 2 2 1 1 2 12 0 1 2 6 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1403:%
149 11 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Calculation of Confidence Intervals.Rnw:ofs 1415:%
1 42 1 1 2 11 0 1 2 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1480:%
162 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Special Cases and Scenarios.Rnw:ofs 1490:%
1 20 1 1 2 4 0 1 2 4 1 1 6 8 0 1 2 4 1 1 6 8 0 1 2 18 1 1 8 16 1 1 7 9 0 1 2 2 1 1 2 8 0 1 2 2 1 1 2 8 0 1 2 21 1 1 3 5 0 1 2 28 1 1 3 5 0 1 2 40 1 1 3 5 0 1 2 25 1 1 3 5 0 1 2 19 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1776:%
173 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Analysis and Visualization.Rnw:ofs 1786:%
1 20 1 1 4 6 0 1 2 3 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 2 1 1 7 9 0 1 2 2 1 1 8 10 0 1 2 4 1 1 5 7 0 1 2 6 1 1 2 4 0 1 2 2 1 1 2 26 0 1 2 16 1 1 4 1 1 1 2 11 0 1 2 18 1 1 5 7 0 1 2 4 1 1 3 5 0 1 2 10 1 1 2 4 0 1 2 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 2017:%
184 8 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Acknowledgements.Rnw:ofs 2026:%
1 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 2031:%
194 11 1}

>>>>

\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:%
1 107 1 1 4 1 1 1 4 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Introduction.Rnw:ofs 115:%
1 97 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 213:%
123 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Main structure and modules of the Package.Rnw:ofs 223:%
1 134 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 358:%
134 11 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Twophase_Estimators and their Application.Rnw:ofs 370:%
1 174 1 1 4 1 1 1 2 1 0 2 1 3 0 1 2 2 1 1 3 14 0 1 2 16 1 1 5 7 0 1 2 4 1 1 2 24 0 1 2 6 1 1 2 4 0 1 2 4 1 1 2 1 0 1 5 6 0 1 2 3 1 1 2 8 0 1 2 129 1 1 7 6 0 1 1 26 0 1 2 3 1 1 2 13 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 7 1 1 6 5 0 1 1 8 0 1 2 7 1 1 2 4 0 1 2 1 1 1 6 5 0 1 1 3 0 1 2 1 1 1 2 11 0 1 2 3 1 1 6 5 0 1 1 8 0 1 2 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 953:%
147}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Threephase_Estimators and their Application.Rnw:ofs 954:%
1 105 1 1 2 1 0 1 1 3 0 1 2 4 1 1 4 1 1 1 6 8 0 1 2 2 1 1 2 27 0 1 2 175 1 1 3 2 0 1 1 3 0 1 2 4 1 1 8 10 0 1 2 1 1 1 4 1 1 1 2 12 0 1 2 4 1 1 7 9 0 1 2 2 1 1 2 12 0 1 2 6 1 1 7 9 0 1 2 2 1 1 2 12 0 1 2 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1398:%
149 11 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Calculation of Confidence Intervals.Rnw:ofs 1410:%
1 42 1 1 2 11 0 1 2 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1475:%
162 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Special Cases and Scenarios.Rnw:ofs 1485:%
1 20 1 1 2 4 0 1 2 4 1 1 6 8 0 1 2 4 1 1 6 8 0 1 2 18 1 1 8 16 1 1 7 9 0 1 2 2 1 1 2 8 0 1 2 2 1 1 2 8 0 1 2 21 1 1 3 5 0 1 2 28 1 1 3 5 0 1 2 40 1 1 3 5 0 1 2 25 1 1 3 5 0 1 2 19 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 1771:%
173 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Analysis and Visualization.Rnw:ofs 1781:%
1 20 1 1 4 6 0 1 2 3 1 1 6 8 0 1 2 2 1 1 7 9 0 1 2 2 1 1 7 9 0 1 2 2 1 1 8 10 0 1 2 4 1 1 5 7 0 1 2 6 1 1 2 4 0 1 2 2 1 1 2 26 0 1 2 16 1 1 4 1 1 1 2 11 0 1 2 18 1 1 5 7 0 1 2 4 1 1 3 5 0 1 2 10 1 1 2 4 0 1 2 9 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 2012:%
184 8 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:./rnw/Acknowledgements.Rnw:ofs 2021:%
1 4 1}
\Sconcordance{concordance:JStatSoft_forestinventory_master.tex:JStatSoft_forestinventory_master.rnw:ofs 2026:%
194 11 1}

### 2017-10-04 rnw/Introduction.Rnw
update Andreas 04.10.2017(1)

Terrestrial inventories have the benefit of being very flexible in the sense that they can be used to produce high quality estimates for a wide-variety of different forest attributes. However, they have the downside of being very expensive.  In order to increase the precision of the estimates, the number of sample plots must be increased, which essentially means travel costs will ballon as trained inventorists are sent to more and more random plot locations. This is why the number of terrestrial samples is usually limited. Whereas in national inventories the terrestrial sample size is still sufficient to provide high estimation accuracies on the national scale, areas with smaller sample sizes, such as forest management units, often are associated with undesirably high estimation errors.  As a result, in recent years there has been increasing need for alternative inventory methods that can maintain the same estimation precision at lower costs, or achieve higher estimation precision at identical costs \citep{vonluepke2013}. A method which has become particularly attractive is called \textit{multiphase sampling}. The core concept is to enlarge the sample size in order to gain higher estimation precision \textit{without} enlarging the terrestrial sample size. This is done by using predictions of the terrestrial target variable at additional sample locations where the terrestrial information has not been gathered. These predictions are produced by regression models that use explanatory variables derived from auxiliary data, commonly in the form of spatially exhaustive remote sensing data in the inventory area. Regression estimators using this concept can consider either \textit{one} additional sample of plot locations consisting of auxiliary information (two-phase or double-sampling) or \textit{two} additional samples of auxiliary information available in different sample sizes (three-phase or triple-sampling) \citep{gregoire2007, saborowski2010, mandallaz2013a, mandallaz2013c, vonLüpke2012}. Their application to existing forest inventory systems have already showed their efficiency in terms of cost reduction and gain in estimation precision \citep{breidenbach2012, vonLuebke2014, mandallaz2013b, magnussen2014, massey2014a}.\par

>>>>

Terrestrial inventories have the benefit of being very flexible in the sense that they can be used to produce high quality estimates for a wide-variety of different forest attributes. However, they have the downside of being very expensive.  In order to increase the precision of the estimates, the number of sample plots must be increased, which essentially means travel costs will rise as trained inventorists are sent to more and more random plot locations. This is why the number of terrestrial samples is usually limited. Whereas in national inventories the terrestrial sample size is still sufficient to provide high estimation accuracies on the national scale, areas with smaller sample sizes, such as forest management units, often are associated with undesirably high estimation errors.  As a result, in recent years there has been increasing need for alternative inventory methods that can maintain the same estimation precision at lower costs, or achieve higher estimation precision at identical costs \citep{vonluepke2013}. A method which has become particularly attractive is called \textit{multiphase sampling}. The core concept is to enlarge the sample size in order to gain higher estimation precision \textit{without} enlarging the terrestrial sample size. This is done by using predictions of the terrestrial target variable at additional sample locations where the terrestrial information has not been gathered. These predictions are produced by regression models that use explanatory variables derived from auxiliary data, commonly in the form of remote sensing data available in the inventory area in large sample or even wall-to-wall. Regression estimators using this concept can consider either \textit{one} additional sample of plot locations consisting of auxiliary information (two-phase or double-sampling) or \textit{two} additional samples of auxiliary information available in different sample sizes (three-phase or triple-sampling) \citep{gregoire2007, saborowski2010, mandallaz2013a, mandallaz2013c, vonLüpke2012}. Their application to existing forest inventory systems have already showed their efficiency in terms of cost reduction and gain in estimation precision \citep{breidenbach2012, vonLuebke2014, mandallaz2013b, magnussen2014, massey2014a}.\par

### 2017-10-04 rnw/Special Cases and Scenarios.Rnw
update Andreas 04.10.2017(1)

If a categorical variable is used in the regression model(s) and the terrestrial sample $s_2$ is considerably small compared to the $s_1$ phase, it might occur that a category is only present in the $s_1 \not\in s_2$ sample, and thus missing in the $s_2$ sample. Conclusively, an internal regression model can not be calculated and the function \textbf{stops} with the following error message:

>>>>

If a categorical variable is used in the regression model(s) and the terrestrial sample $s_2$ is considerably small compared to the $s_1$ phase, it might occur that a category level is only present in the $s_1 \backslash s_2$ sample, and thus missing in the $s_2$ sample. Conclusively, an internal regression model can not be calculated and the function \textbf{stops} with the following error message:

### 2017-10-04 rnw/Threephase_Estimators and their Application.Rnw
update Andreas 04.10.2017(1)

We see that the \code{threephase()}-function returns the sample sizes in the entire inventory area as well as within each small area. The value \code{Inf} for \code{n0G} indicates that the explanatory variables at $s_0$ sample locations used in the reduced model were in our case derived exhaustively. If we compare the two results, we see that the synthetic estimation again yields a much smaller variance than the extended synthetic estimation, but to the cost of ignoring the model uncertainties.

@@

% Finally note that in practice, there are basically \textit{three scenarios} how a two-phase sample design might be extended to a three-phase estimation procedure. In the first scenario, more precise auxiliary information becomes available at a subset of the allready existing and used sample locations. In this case, one simply reclassifies the existing sample locations $x \in s_1$ into $s_0$ sample locations and assigns the $s_1$ indicator to those points where the new variables are available. In the second scenario, the auxiliary information is available at all sample locations $x \in s_1$, but the computation of some of the explanatory variables is computationally expensive. In this case, a considerable time saving could be realized by restricing their computation to a subset of all $s_1$ sample locations, which consequently leads to a three-phase scenario that is handeled likewise the first scenario. A respective example is given in \citet{mandallaz2013b} and \citet{mandallaz2013c}. In both the first and second scenario, the original sample size of the two-phase inventory remains unchanged, but the $s_1$ sample locations are splitted into $s_0$ and $s_1$ locations. In the third and last scenario, one has already used a set of explanatory variables at $s_1$ sample locations for a time before additional, less precise information becomes available in a much larger quantity over the inventory area. It still might yield an increase in estimation precision if this information is considered in the estimation in a much larger sample size than the more precise information (sample fraction). In this case, one creates a new phase $s_0$ by densifying the allready existing $s_1$ sample locations (which in practice are often arranged in regular grids) such that $n_0 >> n_1$ and $s_1 \in s_0$.

>>>>

We see that the \code{threephase()}-function returns the sample sizes in the entire inventory area as well as within each small area. The value \code{Inf} for \code{n0G} indicates that the explanatory variables at $s_0$ sample locations used in the reduced model were in our case derived exhaustively. If we compare the two results, we see that the synthetic estimation again yields a much smaller variance than the extended synthetic estimation, but at the cost of ignoring the model uncertainties.

@@

### 2017-10-04 rnw/Twophase_Estimators and their Application.Rnw
update Andreas 04.10.2017(1)

\begin{figure}[htb]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
    \resizebox{1.12\hsize}{!}{\includegraphics{fig/exh_and_boundweights(2).PNG}}%
		\caption{} \label{fig:exh_nexh_and_boundweights_a}
		\end{subfigure}
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
   \resizebox{0.77\hsize}{!}{\includegraphics{fig/nexh_and_boundweights.PNG}}%
		\caption{} \label{fig:exh_nexh_and_boundweights_b}
	\end{subfigure}
\caption{Concept of exhaustive, see \textbf{(a)}, and non-exhaustive, see \textbf{(b)}, calculation of explanatory variables including boundary adjustment at the support level. Auxiliary data are in both cases available over the entire inventory area marked by the large rectangle. A set $\pmb{Z}$ of $p$ explanatory variables $x_{1},...,x_{p}$ is calculated within the small squares at each sample location $x$ (points) that falls into the forest area (green underlying polygon). w: weights used for calculating the weighted mean of each explanatory variable. Dotted circles indicate terrestrial sample plots.}
\label{fig:exh_nexh_bw}
\end{figure}

@@

\begin{subequations}\label{eq:pest_2p_synth_psynth}
\begin{align}
  \hat{Y}_{G,synth,2p} & =\bar{\pmb{Z}}_G^t\hat{\pmb{\beta}}_{s_2} \label{eq:pointest_2p_reg_synth} \\
  \hat{Y}_{G,psynth,2p} & =\hat{\bar{\pmb{Z}}}_G^t\hat{\pmb{\beta}}_{s_2} \label{eq:pointest_2p_reg_psynth} \\
% \end{align}
% \end{subequations}
% \begin{subequations}\label{eq:var_2p_synth_psynth}
% \begin{align}
  \hat{\var}(\hat{Y}_{G,synth,2p}) & = \bar{\pmb{Z}}_G^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\bar{\pmb{Z}}_G \label{eq:var_2p_reg_synth} \\
  \hat{\var}(\hat{Y}_{G,psynth,2p}) & = \hat{\bar{\pmb{Z}}}_G^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\hat{\bar{\pmb{Z}}}_G
  + \hat{\pmb{\beta}}_{s_2}^t\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}_G}\hat{\pmb{\beta}}_{s_2}  \label{eq:var_2p_reg_psynth}
\end{align}
\end{subequations}

@@

The synthetic estimators, \textit{synth} and \textit{psynth}, have attractively compact formulas but come with the downside of potential bias in their point estimates as well as unreliably small variances. The \textit{small} and \textit{psmall} estimators overcome this issue by using a bias correction term, i.e. $\frac{1}{n_{2,G}}\sum_{x \in s_{2,G}}\hat{R}(x)$. The motivation behind the \textbf{\textit{extended}} \textbf{synthetic} and \textbf{\textit{extended}} \textbf{pseudo synthetic estimator} (\textit{extsynth} and \textit{extpsynth}) is to use the same mathematically elegant formulas of the (pseudo) synthetic estimators while at the same time ensuring that the prediction model residuals over the entire area $F$ and the small area $G$ are by definition both zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This is accomplished by extending the vector of auxiliary information $\pmb{Z}(x)$ by a binary categorical indicator variable $I_G(x)$ which takes the value 1 if the sample location $x$ lies inside the target small area $G$ and is otherwise set to 0. Recalling that linear models fit using OLS have zero mean residuals property by construction, this leads to unbiased point estimates and is also true if categorical variables are included in the model. The new \textit{extended} auxiliary vector thus becomes $\pmb{\mathbb{Z}}^t(x)=(\pmb{Z}^t(x),I_G(x))$ and can be used to replace its non-extended counterpart $\pmb{Z}^t(x)$ whereever it is used in Equations \ref{eq:pest_2p_synth_psynth}, \ref{eq:var_2p_synth_psynth} and  \ref{estvarcovaux_G}. Note that the package functions internally extent the data set by the indicator variable if the \textit{extsynth} or \textit{extpsynth} estimator is called.

>>>>

\begin{figure}[htb]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
    %\resizebox{1.12\hsize}{!}{\includegraphics{fig/exh_and_boundweights(2).PNG}}%
    \resizebox{0.85\hsize}{!}{\includegraphics{fig/boundaryweight_graphic(left).eps}}%
		\caption{} \label{fig:exh_nexh_and_boundweights_a}
		\end{subfigure}
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
   %\resizebox{0.77\hsize}{!}{\includegraphics{fig/nexh_and_boundweights.PNG}}%
   \resizebox{0.85\hsize}{!}{\includegraphics{fig/boundaryweight_graphic(right).eps}}%
		\caption{} \label{fig:exh_nexh_and_boundweights_b}
	\end{subfigure}
\caption{Concept of exhaustive, see \textbf{(a)}, and non-exhaustive, see \textbf{(b)}, calculation of explanatory variables including boundary adjustment at the support level. Auxiliary data are in both cases available over the entire inventory area marked by the large rectangle. A vector of explanatory variables $\pmb{Z}(x)$ is calculated within the supports (small squares)  at each sample location $x$ (points) that falls into the forest area (green underlying polygon). w: weights used for calculating the weighted mean of each explanatory variable. Dotted circles indicate terrestrial sample plots.}
\label{fig:exh_nexh_bw}
\end{figure}

@@

\begin{subequations}\label{eq:pest_2p_synth_psynth}
\begin{align}
  \hat{Y}_{G,synth,2p} & =\bar{\pmb{Z}}_G^t\hat{\pmb{\beta}}_{s_2} \label{eq:pointest_2p_reg_synth} \\
  \hat{Y}_{G,psynth,2p} & =\hat{\bar{\pmb{Z}}}_G^t\hat{\pmb{\beta}}_{s_2} \label{eq:pointest_2p_reg_psynth} \\
  \hat{\var}(\hat{Y}_{G,synth,2p}) & = \bar{\pmb{Z}}_G^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\bar{\pmb{Z}}_G \label{eq:var_2p_reg_synth} \\
  \hat{\var}(\hat{Y}_{G,psynth,2p}) & = \hat{\bar{\pmb{Z}}}_G^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\hat{\bar{\pmb{Z}}}_G
  + \hat{\pmb{\beta}}_{s_2}^t\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}_G}\hat{\pmb{\beta}}_{s_2}  \label{eq:var_2p_reg_psynth}
\end{align}
\end{subequations}

@@

The synthetic estimators, \textit{synth} and \textit{psynth}, have attractively compact formulas but come with the downside of potential bias in their point estimates as well as unreliably small variances. The \textit{small} and \textit{psmall} estimators overcome this issue by using a bias correction term, i.e. $\frac{1}{n_{2,G}}\sum_{x \in s_{2,G}}\hat{R}(x)$. The motivation behind the \textbf{\textit{extended}} \textbf{synthetic} and \textbf{\textit{extended}} \textbf{pseudo synthetic estimator} (\textit{extsynth} and \textit{extpsynth}) is to use the same mathematically elegant formulas of the (pseudo) synthetic estimators while at the same time ensuring that the prediction model residuals over the entire area $F$ and the small area $G$ are by definition both zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This is accomplished by extending the vector of auxiliary information $\pmb{Z}(x)$ by a binary categorical indicator variable $I_G(x)$ which takes the value 1 if the sample location $x$ lies inside the target small area $G$ and is otherwise set to 0. Recalling that linear models fit using OLS have zero mean residuals property by construction, this leads to unbiased point estimates and is also true if categorical variables are included in the model. The new \textit{extended} auxiliary vector thus becomes $\pmb{\mathbb{Z}}^t(x)=(\pmb{Z}^t(x),I_G(x))$ and can be used to replace its non-extended counterpart $\pmb{Z}^t(x)$ whereever it is used in Equations \ref{eq:pest_2p_synth_psynth} and \ref{estvarcovaux_G}. Note that the package functions internally extent the data set by the indicator variable if the \textit{extsynth} or \textit{extpsynth} estimator is called.
